---
title: "The Gamma-Poisson Model"
author: "Dr. Dogucu"
output: 
  xaringan::moon_reader:
    css: slide-style.css
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
---

layout: true
  
<div class="my-header"></div>

<div class="my-footer"> From Bayes Rules! book
 Copyright &copy; Drs. Alicia Johnson, Miles Ott & Mine Dogucu. <a href="https://creativecommons.org/licenses/by-nc-sa/4.0/">CC BY-NC-SA 4.0</a></div> 

---

class: middle

## Choosing a Prior

Prior distribution depends on our current information. When choosing a prior we may consider 

- computational ease
- interpretability

---

## Conjugate Priors

Let $f(\theta)$ be a prior distribution for parameter $\theta$ from a particular distribution, and $L(\theta|x)$ be the likelihood function.
If the resulting posterior distribution $f(\theta|x)$ is of the same family of distributions as the prior distribution, then the prior distribution is a conjugate prior for this likelihood.

--

__Examples__

The Beta-Binomial Model - Beta is a conjugate prior for the Binomial Likelihood.

--

The Gamma-Poisson Model

--

The Normal-Normal Model

---

## Non-Conjugate Prior for Binomial Likelihood

$$f(\pi)=e-e^\pi\; \text{ for } \pi \in [0,1] $$

- Is this a valid pdf?

- After observing 10 successes in 50 trials, calculate the posterior.


---

## The Question

We are interested in modeling $\lambda$ the rate of fraud risk calls received per day. We plan on collecting data on the number of fraud risk phone calls received each day.

--

$\lambda$ is a rate. $\lambda \geq 0$

---

## Gamma Prior

We guess that we receive on average 3 calls per day and $\lambda$ is somewhere between 2 and 4.

--

Let $\lambda$ be a random variable which can take any value between 0 and $\infty$, ie. $\lambda \in [0,\infty)$.  Then the variability in $\lambda$ might be well modeled by a Gamma model with __shape parameter__ $\alpha > 0$ and __rate parameter__ 
$\beta > 0$: 

$$\lambda \sim \text{Gamma}(\alpha, \beta)$$

--

The Gamma distribution is specified by continuous pdf

$$f(\lambda) = \frac{\beta^\alpha}{\Gamma(\alpha)} \lambda^{\alpha-1} e^{-\beta\lambda} \;\; \text{ for } \lambda \in [0,\infty)$$
---

## Gamma Distribution

What is $f(\lambda)$ if $\lambda =1$ and $\lambda \sim \text{Gamma}(2, 5)$ ?

```{r eval = FALSE}
plot_gamma(shape = 2, rate = 5)
```

--

.pull-left[
```{r echo = FALSE, fig.align='center', fig.width=5, fig.height=4}
library(bayesrules)
library(ggplot2)
plot_gamma(2,5) +
    geom_point(aes(x = 1, y = dgamma(1, 
                                     shape =2, 
                                     rate =5)), colour="blue") +
  geom_text(aes(label= "(1, ?)", x = 1.25, y = 0.25), 
            color = "blue")
```

]



.pull-right[
```{r}
dgamma(x = 1, shape = 2, rate = 5)
```


]


---

class: middle

## Descriptives of Gamma

$E(\lambda) = \frac{\alpha}{\beta}$

$\text{Mode}(\lambda) = \frac{\alpha - 1}{\beta} \text{ where } \alpha\ge1$ 

$\text{Var}(\lambda) = \frac{\alpha}{\beta^2}$

---

## Gamma Tuning

```{r echo=FALSE, gamma-tuning, fig.width = 8, fig.height = 6, message = FALSE, fig.align='center'}
library(tidyverse)
# Set up gamma data
alpha <- c(2,2,2,   3,3,3,   5,5,5)
beta  <- c(2,4,6,   2,4,6,  2,4,6)
gammas <- data.frame(setting = factor(rep(1:9, each = 500)), 
                    x = rep(seq(0, 3, length = 500), 9),
                    alpha = rep(alpha, each = 500), beta = rep(beta, each = 500))
gammas <- gammas %>% 
    mutate(y = dgamma(x, shape = alpha, rate = beta))
levels(gammas$setting) <- paste0("Gamma(",alpha,",",beta,")")

  
g <- ggplot(gammas, aes(x = x, y = y)) + 
  lims(x = c(0,1), y = c(0,2.5)) + 
  geom_line() + 
  facet_wrap(~ setting) + 
  labs(x = expression(lambda), y = expression(paste("f(",lambda,")"))) +   scale_x_continuous(breaks = c(0,1,2,3), labels = c("0","1","2","3"))

g
``` 

---

## Tuning Gamma example

For our example on spam phone calls, set a prior for $\lambda$ such that $E(\lambda) = 3$ and $\lambda$ most likely is between 2 and 4. You can use `plot_gamma()` function to try out different gamma distributions.

--

```{r fig.width = 4, fig.height = 3, message = FALSE, fig.align='center'}
plot_gamma(12,4)
```

---

## Poisson Likelihood

We collect four days of data and receive 6, 2, 2, 1 spam calls each day. Write out the likelihood model.

--

$$L(\lambda | (x_1, \ldots, x_n))=\frac{e^{-n\lambda}\lambda^{\sum x_i}}{\prod x_i!}$$

---

## Gamma- Poisson Conjugacy

Based on the observed data how does our idea about $\lambda$ change? Find the posterior.

--

If $f(\lambda) \sim  \text{Gamma}(\alpha, \beta)$   

and if $x_i \sim iid \text{ Poissson}(\lambda) \text{for } i \in 1,\ldots ,n$ 

then $f(\lambda|\vec{x}) \sim \text{Gamma}(\alpha+ \sum x_i, \beta+n)$.


