---
title: "The Beta-Binomial Model"
author: "Dr. Dogucu"
output: 
  xaringan::moon_reader:
    css: slide-style.css
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
---

layout: true
  
<div class="my-header"></div>

<div class="my-footer"> From Bayes Rules! book
 Copyright &copy; Drs. Alicia Johnson, Miles Ott & Mine Dogucu. <a href="https://creativecommons.org/licenses/by-nc-sa/4.0/">CC BY-NC-SA 4.0</a></div> 

---

## Posterior for the Beta-Binomial Model

Let $\pi \sim \text{Beta}(\alpha, \beta)$ and $X|n \sim \text{Bin}(n,\pi)$. 

--

$f(\pi|x) \propto \frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\pi^{\alpha-1} (1-\pi)^{\beta-1} {n \choose x}\pi^x(1-\pi)^{n-x}$

--


$f(\pi|x) \propto \pi^{(\alpha+x)-1} (1-\pi)^{(\beta+n-x)-1}$

--

$\pi|x \sim \text{Beta}(\alpha +x, \beta+n-x)$

--

$f(\pi|x) = \frac{\Gamma(\alpha+\beta+n)}{\Gamma(\alpha+x)\Gamma(\beta+n-x)} \pi^{(\alpha+x)-1} (1-\pi)^{(\beta+n-x)-1}$

---
## Conjugate prior

We say that $f(\pi)$ is a conjugate prior for $L(\pi|x)$ if the posterior, $f(\pi|x) \propto f(\pi)L(\pi|x)$, is from the same model family as the prior.  

Thus, Beta distribution is a conjugate prior for the Binomial likelihood model since the posterior also follows a Beta distribution.

---
## Bike ownership
The transportation office at our school wants to know $\pi$ the proportion of people who own bikes on campus so that they can build bike racks accordingly. The staff at the office think that the $\pi$ is more likely to be somewhere 0.05 to 0.25. The plot below shows their prior distribution. Write out a reasonable $f(\pi)$. Calculate the prior expected value, mode, and variance. 

```{r echo = FALSE}
library(ggplot2)
library(bayesrules)
```


```{r fig.align='center', fig.height=3.5, echo = FALSE}
plot_beta(5, 35) +
    theme(text = element_text(size=20)) 
```

---

## Plotting the prior

```{r eval = FALSE}
plot_beta(5, 35) 
```

```{r fig.align='center', fig.height=4, echo = FALSE}
plot_beta(5, 35) +
    theme(text = element_text(size=20)) 
```

---

## Summarizing the prior

```{r}
summarize_beta(5, 35)
```

---
## Bike ownership posterior

The transportation office decides to collect some data and samples 50 people on campus and asks them whether they own a bike or not. It turns out that 25 of them do! What is the posterior distribution of $\pi$ after having observed this data? 

--

$\pi|x \sim \text{Beta}(\alpha +x, \beta+n-x)$

--

$\pi|x \sim \text{Beta}(5 +25, 35+50-25)$

--

$\pi|x \sim \text{Beta}(30, 60)$


---

## Plotting the posterior

```{r eval = FALSE}
plot_beta(30, 60) 
```

```{r fig.align='center', fig.height=4, echo = FALSE}
plot_beta(30, 60) +
    theme(text = element_text(size=20)) 
```


---

## Summarizing the posterior

```{r}
summarize_beta(30,60)
```



---

## Plot summary

```{r eval = FALSE}
plot_beta(30, 60, mean = TRUE, mode = TRUE) 
```

```{r fig.align='center', fig.height=4, echo = FALSE}
plot_beta(30, 60, mean = TRUE, mode = TRUE) +
    theme(text = element_text(size=20)) 
```

---

## Bike ownership: balancing act

```{r eval = FALSE}
plot_beta_binomial(alpha = 5, beta = 35,
                   x = 25, n = 50)
```

```{r fig.align='center', fig.height=5, echo = FALSE}
plot_beta_binomial(alpha = 5, beta = 35,
                   x = 25, n = 50) +
      theme(text = element_text(size=20)) 

```



---

## Posterior Descriptives

$\pi|(X=x) \sim \text{Beta}(\alpha+x, \beta+n-x)$

$$E(\pi | (X=x)) = \frac{\alpha + x}{\alpha + \beta + n}$$ 
$$\text{Mode}(\pi | (X=x))  = \frac{\alpha + x - 1}{\alpha + \beta + n - 2} $$
$$\text{Var}(\pi | (X=x))   = \frac{(\alpha + x)(\beta + n - x)}{(\alpha + \beta + n)^2(\alpha + \beta + n + 1)}\\$$
---

## Bike ownership - descriptives of the posterior

What is the descriptive measures (expected value, mode, and variance) of the posterior distribution for the bike ownership example?

--


```{r}
summarize_beta_binomial(5, 35, x = 25, n = 50)
```

